---
title: "Exploratoy Data Analysis in R"
format: html
editor: visual
---

## EDA for Jordan Retail Dataset

Open the dataset (Read Excel File)

```{r}
install.packages("readxl")

library(readxl)

file_path <- 'G:\\My Drive\\Current\\Datasets\\jordan_retail_single_product_orders.xlsx'

data <- read_excel(file_path, sheet = 1)
```

View the data in the interactive window

```{r}
View(data)
```

Show the datatype of the mtcars variable

```{r}
class(data)
```

\
You can perform various descriptive statistics on your sales data in R to gain insights into the data. Here are some common descriptive statistics you can calculate:

1.  Summary Statistics:

    -   **`summary(data)`** to get summary statistics for all numerical variables.

    -   **`summary(data$unit_price)`** to get summary statistics for the unit price.

2.  Mean and Median:

    -   **`mean(data$unit_price)`** to calculate the mean unit price.

    -   **`median(data$unit_price)`** to calculate the median unit price.

3.  Variance and Standard Deviation:

    -   **`var(data$unit_price)`** to calculate the variance of unit price.

    -   **`sd(data$unit_price)`** to calculate the standard deviation of unit price.

4.  Count and Frequency:

    -   **`table(data$product_line)`** to count the frequency of each product line.

5.  Histogram:

    -   **`hist(data$unit_price)`** to create a histogram of unit prices.

6.  Box Plot:

    -   **`boxplot(data$unit_price)`** to create a box plot of unit prices.

7.  Correlation:

    -   **`cor(data$unit_price, data$quantity)`** to calculate the correlation between unit price and quantity.

8.  Cross-tabulation:

    -   **`table(data$customer_type, data$customer_gender)`** to create a cross-tabulation between customer type and gender.

9.  Percentiles:

    -   **`quantile(data$unit_price, c(0.25, 0.50, 0.75))`** to calculate quartiles of unit prices.

10. Frequency Tables:

    -   **`table(data$branch_name)`** to create a frequency table for branch names.

    -   **`table(data$payment_method)`** to create a frequency table for payment methods.

Display summary statisitcs

```{r}
summary(data)
```

## Possible Explorations

## EDA using Visualization

### Histograms

```{r}
#create histogram of values for mpg
hist(mtcars$mpg,
     col='steelblue',
     main='Car Distribution according to Miles/(US) gallon',
     xlab='Miles/(US) gallon',
     ylab='Frequency')
```

### Box Plots

```{r}
#create boxplot of values for mpg
boxplot(mtcars$mpg,
        main='Distribution of Miles/(US) gallon values',
        ylab='Miles/(US) gallon',
        col='steelblue',
        border='black')
```

#### Plot Analysis

This plot demonstrates the distribution of the mpg values for the 999 cars in the dataset. We present below the observations//analysis of the plot.

-   The the mpg values range spans from 10 to 40 while the majority of observations are centered around 15 to 23 mpg.
-   We can see that the majority of cars have a value close to 20 mpg.
-   We can also notice a slight skewness towards the rigth since we can see a longer right tail.
-   We do not observe any outliers in our data.

### Scatter Plots

We can also use the plot() function to create a scatterplot of any pairwise combination of variables:

```{r}
#create scatterplot of mpg vs. wt
plot(mtcars$mpg, mtcars$wt,
     col='steelblue',
     main='Scatterplot',
     xlab='Miles/(US) gallon',
     ylab='Weight (1000 lbs)',
     pch=19)
```

```{r}

```

### Five Number Summary
