---
title: "Initial Exploratoy Data Analysis in R"
format: html
editor: visual
---

# Initial EDA using mtcars Dataset

\
The mtcars dataset is **a built-in dataset in R**. It comprises 11 features of 32 automobiles from the 1974 Motor Trend US magazine. We will use it for this data exploration exercise.

View the mtcars in the interactive window

```{r}
# View(mtcars)
```

Show the datatype of the mtcars variable

```{r}
class(mtcars)
```

Display help about the dataset

```{r}
?mtcars
```

### Display summary statistics

```{r}
summary(mtcars)
```

## Initial EDA using Visualization

### Histograms

```{r}
#create histogram of values for mpg
hist(mtcars$mpg,
     col='steelblue',
     main='Car Distribution according to Miles/(US) gallon',
     xlab='Miles/(US) gallon',
     ylab='Frequency')
```

### Box Plots

```{r}
#create boxplot of values for mpg
boxplot(mtcars$mpg,
        main='Distribution of Miles/(US) gallon values',
        ylab='Miles/(US) gallon',
        col='steelblue',
        border='black')
```

#### Plot Analysis

This plot demonstrates the distribution of the mpg values for the 999 cars in the dataset. We present below the observations//analysis of the plot.

-   The the mpg values range spans from 10 to 40 while the majority of observations are centered around 15 to 23 mpg.
-   We can see that the majority of cars have a value close to 20 mpg.
-   We can also notice a slight skewness towards the rigth since we can see a longer right tail.
-   We do not observe any outliers in our data.

### Scatter Plots

We can also use the plot() function to create a scatterplot of any pairwise combination of variables:

```{r}
#create scatterplot of mpg vs. wt
plot(mtcars$mpg, mtcars$wt,
     col='steelblue',
     main='Scatterplot',
     xlab='Miles/(US) gallon',
     ylab='Weight (1000 lbs)',
     pch=19)
```

```{r}

```

### Five Number Summary
