---
title: "Anova Test"
format: html
editor: visual
---

Certainly! To conduct an ANOVA (Analysis of Variance) test in R, you'll first need some data where you are interested in comparing the means of a numerical variable across multiple groups defined by a categorical variable. For this example, let's create a hypothetical dataset where we compare the scores of students across three different majors.

### Step 1: Create a Hypothetical Dataset

```{r}
# Create a data frame with scores and corresponding student majors
set.seed(123)  # for reproducible results
data <- data.frame(
  score = c(rnorm(30, mean = 75, sd = 10),  # scores for major A
            rnorm(30, mean = 80, sd = 10),  # scores for major B
            rnorm(30, mean = 85, sd = 10)), # scores for major C
  major = factor(rep(c("A", "B", "C"), each = 30))  # corresponding majors
)
```

In this dataset, we have three majors (A, B, C) and scores that follow a normal distribution with different means for demonstration purposes.

### Step 2: Conduct ANOVA Test

We'll use the `aov()` function to perform the ANOVA test, testing the null hypothesis that all group means are equal against the alternative that at least one group mean is different.

```{r}
# Conduct the ANOVA test
anova_result <- aov(score ~ major, data = data)
summary(anova_result)
```

### Step 3: Interpret the Results

The `summary()` function on the ANOVA result object will provide you with a table showing the ANOVA summary, including the F-statistic and the p-value.

-   The **F-statistic** is used to determine whether there are any statistically significant differences between the means of the groups.
-   The **p-value** tells you whether the observed data could occur by chance if the null hypothesis were true. A small p-value (typically ≤ 0.05) indicates strong evidence against the null hypothesis, so you reject the null hypothesis.

### Example Interpretation:

If the p-value is less than 0.05, you might conclude that there is a statistically significant difference in scores among the different majors.

Please note, ANOVA tells you if there's a significant difference but doesn't tell you where that difference lies. To determine which specific groups differ from each other, you would follow up with a post hoc test, such as Tukey's HSD (Honest Significant Difference) test.

### Post Hoc Analysis Example (if ANOVA is significant):

```{r}

summary(anova_result)
anova_result

if (summary(anova_result)[["Pr(>F)"]][1] < 0.05) {
  posthoc_result <- TukeyHSD(anova_result)
  print(posthoc_result)
 }
```

My ChatGPT This code snippet checks if the ANOVA result is significant, and if so, it proceeds to perform Tukey's HSD test to identify which groups' means are significantly different from each other.

User Here is the result for the code above, what does it tell us?

```         
        Df Sum Sq Mean Sq F value   Pr(>F)    
```

major 2 1292 646.2 31.83 4.23e-11 ***Residuals 87 1766 20.3\
--- Signif. codes: 0 ‘***’ 0.001 ‘\*\*’ 0.01 ‘\*’ 0.05 ‘.’ 0.1 ‘ ’ 1

The output you've provided is from the ANOVA (Analysis of Variance) summary table for comparing the means of a numerical variable (scores) across three different groups defined by a categorical variable (majors) in R. Here's how to interpret the key components of this output:

### Components of the ANOVA Table

-   **Df (Degrees of Freedom):**
    -   Major: 2, which corresponds to the number of groups (majors) minus 1. This is because you have three groups (A, B, C).
    -   Residuals: 87, calculated as the total number of observations minus the number of groups. If you had 90 observations in total, this makes sense (90 - 3 = 87).
-   **Sum Sq (Sum of Squares):**
    -   Major: 1292, which represents the between-group variability (i.e., the variation in scores attributable to differences between the majors).
    -   Residuals: 1766, which represents the within-group variability (i.e., the variation in scores within each major group).
-   **Mean Sq (Mean Square):**
    -   Major: 646.2, which is the between-group Sum of Squares divided by its degrees of freedom (1292 / 2).
    -   Residuals: 20.3, which is the within-group Sum of Squares divided by its degrees of freedom (1766 / 87).
-   **F value:**
    -   31.83, which is the ratio of the Mean Square between groups to the Mean Square within groups (646.2 / 20.3). This statistic is used to determine whether the group means are significantly different.
-   **Pr(\>F) (p-value):**
    -   4.23e-11, which indicates the probability of observing the data (or something more extreme) if the null hypothesis of no difference between group means is true.

### Interpretation

-   The **p-value is extremely small** (4.23e-11), far below the common significance level of 0.05. This means there is very strong evidence against the null hypothesis, which posits that there are no differences among the group means.
-   Given the p-value, you reject the null hypothesis and conclude that there are statistically significant differences in scores among the different majors.
-   The **F value** is high (31.83), indicating that the ratio of between-group variance to within-group variance is large and that the group means are not all the same.

### Significance Codes

-   The `***` next to the p-value indicates a very high level of statistical significance.

### What's Next

Since the ANOVA test shows significant differences among the group means, you would typically proceed with a post hoc test (such as Tukey's HSD) to pinpoint exactly which groups differ from each other.

### Conclusion

The result of this ANOVA test tells us that the mean scores significantly differ across the three majors. This is a crucial finding for further investigation, possibly exploring why these differences exist and which specific majors are different from each other through post hoc analysis.
