Certainly! Let's use some common sample datasets available in R and demonstrate each of the basic plotting functions with brief examples:

1.  **`plot()`** - Generic function for plotting, typically used for scatter plots.

```{r}
# Using the cars dataset
plot(cars$speed, cars$dist, main = "Stopping Distance vs Speed",
     xlab = "Speed", ylab = "Stopping Distance")
```

2.  **`hist()`** - For creating histograms.

```{r}
# Using the faithful dataset
hist(faithful$eruptions, breaks = 20, main = "Histogram of Eruptions",
     xlab = "Duration of Eruption (minutes)")
```

3.  **`boxplot()`** - Produces box plots.

```{r}
# Using the iris dataset
boxplot(iris$Sepal.Length ~ iris$Species, main = "Boxplot of Sepal Length by Species",
        ylab = "Sepal Length (cm)", xlab = "Species")
```

4.  **`barplot()`** - Creates bar plots.

```{r}
# Using a table from the VADeaths dataset
death_rates <- VADeaths[1,]
barplot(death_rates, main = "Death Rates in Virginia (1940)",
        ylab = "Rate (per 1000)", xlab = "Age Group")
```

5.  **`pie()`** - Generates pie charts.

```{r}
# Using a simple named vector
slices <- c(10, 12, 4, 16, 8)
lbls <- c("US", "UK", "Australia", "Germany", "France")
pie(slices, labels = lbls, main = "Pie Chart of Countries")
```

6.  **`stripchart()`** - Creates strip charts.

```{r}
# Using the iris dataset
stripchart(iris$Sepal.Length ~ iris$Species, method = "jitter",
           main = "Stripchart of Sepal Length by Species",
           xlab = "Species", ylab = "Sepal Length (cm)")
```

7.  **`dotchart()`** - Produces dot plots.

```{r}
# Using the mtcars dataset
dotchart(mtcars$mpg, labels = row.names(mtcars), main = "Dotchart of MPG", xlab = "Miles Per Gallon")
```

8.  **`pairs()`** - Generates a matrix of scatter plots.

```{r}
# Using the first four columns of the iris dataset
pairs(~Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, data = iris,
      main = "Scatterplot Matrix of Iris Data")
```

9.  **`stem()`** - Produces a stem-and-leaf plot.

```{r}
# Using the trees dataset
stem(trees$Height, scale = 2, main = "Stem-and-Leaf Plot of Tree Heights")
```

10. **`qqnorm()`** and **`qqplot()`** - Used for creating Q-Q plots.

```{r}
# qqnorm with the rivers dataset
qqnorm(rivers, main = "Q-Q Plot of Rivers Length")
qqline(rivers)

# qqplot comparing two datasets
qqplot(rivers, faithful$eruptions, main = "Q-Q Plot of Rivers Length vs. Eruptions Duration")
```

Each of these plots leverages one of the built-in datasets (`cars`, `faithful`, `iris`, `VADeaths`, `mtcars`, `trees`, `rivers`) that come with R and are used here to demonstrate different types of standard plots available in base R.

## More About Plot Function

In R, the `plot` function can be customized with various types of plotting styles. The `type` argument to `plot` determines the type of plot that will be drawn. Here are examples of some common `type` options:

1.  **`type = "p"`**: Draws only the points (default).
2.  **`type = "l"`**: Draws only the lines.
3.  **`type = "b"`**: Draws both points and lines.
4.  **`type = "h"`**: Draws vertical lines from the x-axis to the y-value.

Let's use the `pressure` dataset for the demonstration, as it's a simple dataset with two columns:

```{r}
# Point Plot
plot(pressure$temperature, pressure$pressure, 
     type = "p", 
     main = "Point Plot", 
     xlab = "Temperature", 
     ylab = "Pressure")

# Line Plot
plot(pressure$temperature, pressure$pressure, 
     type = "l", 
     main = "Line Plot", 
     xlab = "Temperature", 
     ylab = "Pressure")

# Both Points and Lines
plot(pressure$temperature, pressure$pressure, 
     type = "b", 
     main = "Points and Lines", 
     xlab = "Temperature", 
     ylab = "Pressure")

# High-density Plot
plot(pressure$temperature, pressure$pressure, 
     type = "h", 
     main = "High-density Plot", 
     xlab = "Temperature", 
     ylab = "Pressure")
```

Each of these commands will create a different type of plot:

-   `type = "p"` produces a scatter plot showing just the data points.
-   `type = "l"` creates a line plot, connecting the data points with lines.
-   `type = "b"` displays both points and lines, connecting the points with lines.
-   `type = "h"` creates a high-density plot, where vertical lines are drawn from each point down to the x-axis, giving a sense of density or frequency of the points along the axis.

You can try running these commands in your R console or RStudio to see the different plot types.
